Linear Algebra as a Unifying Theory\\of Wireless Workload Modeling

### 2008-02-02 paper.tex
More description of flow behaviors.

\subsection*{Flow Behavior}

The behavior of a flow may be mapped into a vector space such that the two operations, namely vector addition and scalar multiplication, can be sensibly defined in a way that naturally corresponds with the aggregation of behaviors. This is best clarified with an intentionally simplistic example. Suppose that the behavior of a flow is represented by the number of packets, $p$, the total bytes transmitted, $b$, and the total duration, $d$. This triplet can be naturally embedded in Euclidean space: $\vec{x}=(p,b,d)\in\R^3$. If we have two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, then their ``sum behavior,'' $\vec{x}_1+\vec{x}_2$ has:
\begin{align*}
p_1+p_2 &\text{ packets,}\\
b_1+b_2 &\text{ bytes,}\\
d_1+d_2 &\text{ duration.}
\end{align*}
Their average, i.e. $\frac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.

While this may seem trivial, note that the representation must be chosen carefully. Suppose instead, for example, we represented each flow's behavior by its packet count, $p$, average packet size, $b/p$, and average inter-packet interval, $d/(p-1)$. In this representation, the average of two behaviors has the average average packet size (the two averages are intentional), and the average average inter-packet interval. There lies the rub: this is still the case no matter how many packets are in each flow. Instead, we should more naturally weight the averages by the number of packets in each constituent. This is precisely what the packets-bytes-duration representation accomplishes.

>>>>

\subsection{Flow Behavior}

The behavior of a flow may be mapped into a vector space such that the two operations, namely vector addition and scalar multiplication, can be sensibly defined in a way that naturally corresponds with the aggregation of behaviors. This is best clarified with an intentionally simplistic example. Suppose that the behavior of a flow is represented by the number of packets, $p$, the total bytes transmitted, $b$, and the total duration, $d$. This triplet can be naturally embedded in Euclidean space: $\vec{x}=(p,b,d)\in\R^3$. If we have two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, then their ``sum'' behavior, $\vec{x}_1+\vec{x}_2$ has $p_1+p_2$ packets, $b_1+b_2$ bytes, $d_1+d_2$ duration. Their average, i.e. $\frac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.

%While this may seem trivial, note that the representation must be chosen carefully. Suppose instead, for example, we represented each flow's behavior by its packet count, $p$, average packet size, $b/p$, and average inter-packet interval, $d/(p-1)$. In this representation, the average of two behaviors has the \textit{average} average packet size, and the \textit{average} average inter-packet interval. Therein lies the rub: this is the case no matter how many packets are in each flow; if a flow with thousands of 1500-byte packets is averaged with a single-packet, one-byte flow, the average behavior has... Instead, we should more naturally weight the averages by the number of packets in each constituent. This is precisely what the packets-bytes-duration representation accomplishes.

The {packets-bytes-duration} model representation of flow behavior is clearly quite simplistic. It captures nothing about variability in packet sizes, nor about the ``burstiness'' of packet transmissions. At the opposite extreme, it is possible to embed the exact start time of each flow, and the exact size and transmission times of its packets in a vector space. The difficulty comes in doing so in a manner that makes the summation of vectors correspond naturally and meaningfully to aggregate behavior of flows. Instead, we take an intermediate approach, suggested by the main result of~\cite{Karpinski07:realism}: if the start time and end time of each flow is known, it suffices to know the empirical distribution of packet sizes and inter-packet intervals in order to reproduce network performance characteristics accurately. Accordingly, we represent a flow's behavior by these precise characteristics.

Explicitly, we encode a flow's behavior with four vectors: $\vec{z} \in \R^{d_z}$ is a histogram of packet sizes; $\vec{v} \in \R^{d_v}$ is a histogram of inter-packet intervals; $\vec{\alpha}, \vec{\epsilon} \in \R^{d_t}$ are vectors indicating when the flow begins and ends. To expand on this, $z_i$ (i.e. the $i$th element of $\vec{z}$) is the number of packets in the flow such with $i$ bytes and $d_z$ is the largest possible packet size, which in practice is the maximum transfer unit of the wireless medium---typically 1500 bytes. The histogram of inter-packet intervals is similar, except that domain of possible intervals is a continuum, so we much map the range of possible intervals onto the discrete index values $\left\{1,\dots,d_v\right\}$. In our experiments, we use the following logarithmic quantization function:
\begin{align}
q(v) &= \left\lfloor(d_v-1)\frac{\ln v - \ln v_{\min}}{\ln v_{\max}-\ln v_{\min}}\right\rfloor+1.
\end{align}
with $v_{\min}=10^{-6}$, which is the minimal interval our trace data, and $v_{\max}=600$, which is the total duration of each simulated situation, and hence an upper bound on the inter-packet intervals. We use $d_v=1500$ to match the dimensionally of the packet size representation.

The flow start and end time indicator vectors, $\vec{\alpha}$ and $\vec{\epsilon}$, also require some explanation. Start and end times are simple scalar values. Why not represent them as such? The answer lies in the requirement that behaviors must add meaningfully. If we represent the start time as a scalar, then adding the behavior of two flows produces a (fairly) meaningless value. It's certainly not a start time itself---it may even exceed the start time of any actual flow. Instead, we again quantize the continuum of time: we map the range of possible start times onto the indices $\left\{1,\dots,d_t\right\}$, where $d_t$ is some chosen dimensionality. If the start time is mapped to index, $i$, then $\vec{\alpha}=e_i$, the vector with 1 at index $i$ and zeros elsewhere. The benefit of this representation is that we can now appropriately represent aggregate behavior through vector addition: the sum of start vectors over any collection of flows represents is a histogram of start times for those flows.

Finally, we collect all the pieces of flow behavior into matrix representations of behavior across an entire traffic trace. If there are $f$ flows in the trace, we define the size matrix $Z\in\R^{f \times d_z}$ to be the matrix whose row vectors are the size vectors $\vec{z}_i$ for $i=1,\dots,f$. In precisely the same manner, we define the interval matrix, $V\in\R^{f \times d_v}$, and the start time matrix, $A\in\R^{f \times d_t}$, and the end time matrix, $E\in\R^{f \times d_t}$.

\subsection{Flow Sources \& Destinations}
